---
title: "BioMe_proteome_EDA"
author: "Meizhen Yao"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    fig_caption: yes
    fig_height: 7
    fig_width: 10
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    self_contained: yes
    code_folding: hide
    toc_depth: 6
header-includes: \usepackage{multirow}
editor_options: 
  chunk_output_type: console
---

<style type="text/css">
body{
  /*font-family: Helvetica;*/
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
options(digits = 4)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,cache=F,warning = FALSE)
# suppress warning messages for final rendering
old.warn <- getOption("warn")
options(qwraps2_markup = "markdown")

```

```{r, include=FALSE}
library(tidyverse)
library(ggpubr)
library(data.table)
library(readr)
library(readxl)
library(formattable)
library(ComplexHeatmap)
library(OlinkAnalyze)
library(gtsummary)
library(flextable)
library(ggplotify)
library(pheatmap)
library(plotly)
library(gapminder)
library(manipulateWidget)
library(ggfortify)
library(reactable)
library(mice)
library(naniar)
library(WGCNA)


## import data
BioMe_proteome <- fread("C:/Users/yaom03/OneDrive - The Mount Sinai Hospital/Documents/Projects/BioMe/proteome/input/Q-00771_Loos_NPX_2022-05-03/Q-00771_Loos_NPX_2022-05-03.txt")
assay_list<- read_excel("C:/Users/yaom03/OneDrive - The Mount Sinai Hospital/Documents/Projects/BioMe/proteome/input/assay-list-olink-explore-3072.xlsx", sheet = "data")

## remove controls
BioMe_proteome_nocontrol<- BioMe_proteome %>% 
                           filter(!grepl("CONTROL", SampleID))


## combine data with annotation
str(BioMe_proteome_nocontrol)
str(assay_list)

colnames(assay_list)[1]<- "UniProt"
colnames(assay_list)[2]<- "Protein_name"
colnames(assay_list)[3]<- "Gene_name"
colnames(assay_list)[4]<- "Panel"


BioMe_proteome_protname<- BioMe_proteome_nocontrol %>% 
                          left_join(assay_list, by = c("Panel", "UniProt"))


colSums(is.na(BioMe_proteome_protname))


#------------------------------------------- assign WARN and LOD indicator column
BioMe_proteome_protname<- BioMe_proteome_protname %>% 
                          mutate(Warning = if_else(QC_Warning == "PASS" & Assay_Warning == "PASS", "PASS", "WARN"),
                                 LOD_ind = if_else(NPX >= LOD, 1, 0)) 


#------------------------------------------- transpose to wide format
BioMe_proteome_wide<- BioMe_proteome_protname %>% 
                      select(SampleID, OlinkID, NPX) %>% 
                      pivot_wider(
                        names_from = OlinkID,
                        values_from = NPX
                      )

```


```{r}
sample<- c("352")
protein<- c(2943)
panel<- c(8)

data<- data.frame(sample,
                  protein,
                  panel)
flextable(data)
```

# Data
```{r}

BioMe_proteome_protein_data_table<- BioMe_proteome_protname %>% 
                                    select(Panel, SampleID, OlinkID, Gene_name, Protein_name) %>% 
                                    group_by(Panel) %>% 
                                    as.data.frame() %>% 
                                    reactable(groupBy = c("Panel"))

BioMe_proteome_protein_data_table


```



# Sample exploratory
## Panel & proteins
```{r}

BioMe_proteome_protein_panel_table<- BioMe_proteome_protname %>% 
                                     select(Panel, Protein_name, Gene_name, OlinkID) %>% 
                                     group_by(Panel, Protein_name, Gene_name, OlinkID) %>% 
                                     summarise(Sample_count=n())%>%
                                     as.data.frame() %>% 
                                     reactable(groupBy = c("Panel"))

BioMe_proteome_protein_panel_table


```
 

## Panel & Samples
```{r}

BioMe_proteome_sample_table<- BioMe_proteome_protname %>% 
                                     select(Panel, SampleID) %>% 
                                     group_by(Panel, SampleID) %>% 
                                     arrange(by_group = SampleID) %>% 
                                     summarise(protein_count=n())%>% 
              reactable(groupBy = "Panel") 


BioMe_proteome_sample_table

```

## Duplicate assay
```{r}

BioMe_proteome_protein_dup_table<- BioMe_proteome_protname %>% 
                                   select(Panel, Protein_name) %>% 
                                   distinct() %>% 
                                   group_by(Protein_name) %>% 
                                   summarise(protein_dup=n()) %>% 
                                   reactable(groupBy = "protein_dup")

BioMe_proteome_protein_dup_table


BioMe_proteome_protein_dup_detail<- BioMe_proteome_protname %>% 
                                    select(Panel, Protein_name) %>% 
                                    distinct() %>% 
                                    group_by(Protein_name) %>% 
                                    summarise(protein_dup=n()) %>% 
                                    filter(protein_dup!=1)

dup_details<- BioMe_proteome_protein_dup_detail %>% 
              left_join((BioMe_proteome_protname %>% 
                                   select(Panel, Protein_name) %>% 
                                   distinct()), by="Protein_name") %>% 
              reactable(groupBy = "Protein_name") 
              
dup_details

```




# Data quality
## Checking missing
```{r, warning=FALSE,message=FALSE}

BioMe_proteome_cardio_wide<- BioMe_proteome_protname %>% 
                             filter(Panel == "Cardiometabolic") %>% 
                             select(SampleID, PlateID, OlinkID, NPX) %>% 
                             pivot_wider(
                               names_from = OlinkID,
                               values_from = NPX
                             )

vis_miss(BioMe_proteome_cardio_wide) +
  labs(title="Cardiometabolic")

BioMe_proteome_cardio2_wide<- BioMe_proteome_protname %>% 
                             filter(Panel == "Cardiometabolic_II") %>% 
                             select(SampleID, OlinkID, NPX) %>% 
                             pivot_wider(
                               names_from = OlinkID,
                               values_from = NPX
                             )

vis_miss(BioMe_proteome_cardio2_wide) +
  labs(title="Cardiometabolic_II")


BioMe_proteome_inflam_wide<- BioMe_proteome_protname %>% 
                             filter(Panel == "Inflammation") %>% 
                             select(SampleID, OlinkID, NPX) %>% 
                             pivot_wider(
                               names_from = OlinkID,
                               values_from = NPX
                             )

vis_miss(BioMe_proteome_inflam_wide) +
  labs(title="Inflammation")



BioMe_proteome_inflam2_wide<- BioMe_proteome_protname %>% 
                             filter(Panel == "Inflammation_II") %>% 
                             select(SampleID, OlinkID, NPX) %>% 
                             pivot_wider(
                               names_from = OlinkID,
                               values_from = NPX
                             )

vis_miss(BioMe_proteome_inflam2_wide) +
  labs(title="Inflammation_II")


BioMe_proteome_oncology_wide<- BioMe_proteome_protname %>% 
                             filter(Panel == "Oncology") %>% 
                             select(SampleID, OlinkID, NPX) %>% 
                             pivot_wider(
                               names_from = OlinkID,
                               values_from = NPX
                             )

vis_miss(BioMe_proteome_oncology_wide) +
  labs(title="Oncology")



BioMe_proteome_Oncology2_wide<- BioMe_proteome_protname %>% 
                             filter(Panel == "Oncology_II") %>% 
                             select(SampleID, OlinkID, NPX) %>% 
                             pivot_wider(
                               names_from = OlinkID,
                               values_from = NPX
                             )

vis_miss(BioMe_proteome_Oncology2_wide) +
  labs(title="Oncology_II")


BioMe_proteome_Neurology_wide<- BioMe_proteome_protname %>% 
                             filter(Panel == "Neurology") %>% 
                             select(SampleID, OlinkID, NPX) %>% 
                             pivot_wider(
                               names_from = OlinkID,
                               values_from = NPX
                             )

vis_miss(BioMe_proteome_Neurology_wide) +
  labs(title="Neurology")



BioMe_proteome_Neurology2_wide<- BioMe_proteome_protname %>% 
                             filter(Panel == "Neurology_II") %>% 
                             select(SampleID, OlinkID, NPX) %>% 
                             pivot_wider(
                               names_from = OlinkID,
                               values_from = NPX
                             )

vis_miss(BioMe_proteome_Neurology2_wide) +
  labs(title="Neurology_II")
```


## Checking QC Warning & LOD
### Details for each protein
```{r, warning=FALSE,message=FALSE}

BioMe_proteome_QC_data<- BioMe_proteome_protname %>% 
                          select(Panel, Protein_name, Gene_name, OlinkID, Warning) %>% 
                          group_by(Panel, Protein_name, Gene_name, OlinkID, Warning) %>% 
                          summarise(Warning_num=n())%>%
                          mutate(Warning_sign = case_when(Warning=="PASS" & Warning_num==352 ~ 0, 
                                                          Warning=="PASS" & Warning_num!=352 ~ NA,
                                                          Warning=="WARN" ~ Warning_num)) %>% 
                          filter(is.na(Warning_sign) == FALSE) %>% 
                          mutate(Warning_percent = Warning_sign/352) %>% 
                          select(Panel, Protein_name, Gene_name, OlinkID, Warning_percent) %>% 
                          as.data.frame() 




BioMe_proteome_LOD_data<- BioMe_proteome_protname %>% 
                          select(Panel, Protein_name, Gene_name, OlinkID, LOD_ind) %>% 
                          group_by(Panel, Protein_name, Gene_name, OlinkID, LOD_ind) %>% 
                          summarise(LOD_num=n())%>%
                          mutate(LOD_sign = case_when(LOD_ind==0 & LOD_num==352 ~ 0, 
                                                          LOD_ind==0 & LOD_num!=352 ~ NA,
                                                          LOD_ind==1 ~ LOD_num)) %>% 
                          filter(is.na(LOD_sign) == FALSE) %>% 
                          mutate(Percentage_above_LOD = LOD_sign/352) %>% 
                          select(Panel, Protein_name, Gene_name, OlinkID, Percentage_above_LOD) %>% 
                          as.data.frame() 

BioMe_proteome_check_data<- BioMe_proteome_QC_data %>% 
                            inner_join(BioMe_proteome_LOD_data, by=c('Panel', 'Protein_name', 'Gene_name', 'OlinkID'))


BioMe_proteome_check_table<- BioMe_proteome_check_data %>% 
                             reactable(groupBy = c("Panel"), filterable = TRUE,  
                                    columns = list(Warning_percent = colDef(format = colFormat(percent = TRUE, digits = 1),
                                                                                 style = function(Warning_percent) {
                                                                                         if (Warning_percent > 0.25) {
                                                                                             color <- "#e00000"
                                                                                         } else if (Warning_percent <= 0.25) {
                                                                                             color <- "#008000"
                                                                                         } 
                                                                                           list(color = color, fontWeight = "bold")
                                                                                         }),
                                                   Percentage_above_LOD = colDef(format = colFormat(percent = TRUE, digits = 1),
                                                                                 style = function(Percentage_above_LOD) {
                                                                                         if (Percentage_above_LOD > 0.25) {
                                                                                             color <- "#008000"
                                                                                         } else if (Percentage_above_LOD <= 0.25) {
                                                                                             color <- "#e00000"
                                                                                         } 
                                                                                           list(color = color, fontWeight = "bold")
                                                                                         })))


BioMe_proteome_check_table

BioMe_proteome_warn_num<- BioMe_proteome_check_data %>% 
                           filter(Warning_percent > 0.25) %>%
                           select('Panel', 'Protein_name', 'Gene_name', 'OlinkID', 'Warning_percent') %>% 
                           reactable(groupBy = c("Panel"), 
                                     columns = list(
                                       Warning_percent = colDef(format = colFormat(percent = TRUE, digits = 1)),
                                       Protein_name = colDef(footer = function(values) sprintf("Total = %d", length(values)))))
                           
BioMe_proteome_warn_num                          



BioMe_proteome_LOD_num<- BioMe_proteome_check_data %>% 
                           filter(Percentage_above_LOD < 0.25) %>%
                           select('Panel', 'Protein_name', 'Gene_name', 'OlinkID', 'Percentage_above_LOD') %>% 
                           reactable(groupBy = c("Panel"),
                                     columns = list(
                                       Percentage_above_LOD = colDef(format = colFormat(percent = TRUE, digits = 1)),
                                       Protein_name = colDef(footer = function(values) sprintf("Total = %d", length(values)))))
                           
BioMe_proteome_LOD_num  

BioMe_proteome_remove<- BioMe_proteome_check_data %>% 
                        filter(Warning_percent > 0.25 | Percentage_above_LOD < 0.25)

write.table(BioMe_proteome_remove, "~/Projects/BioMe/proteome/input/analysis_sample/BioMe_proteome_warn_remove.txt", row.names = FALSE)

```

### Details for each sample
```{r, warning=FALSE,message=FALSE}
BioMe_proteome_protname$Warning<- factor(BioMe_proteome_protname$Warning, 
                                         levels = c("PASS", "WARN"))
BioMe_proteome_protname$LOD_ind<- factor(BioMe_proteome_protname$LOD_ind, 
                                         levels = c(1, 0))

BioMe_sample_QC_data<- BioMe_proteome_protname  %>% 
                          group_by(SampleID) %>% 
                          summarise(N = n()) %>% 
                          inner_join((BioMe_proteome_protname %>% 
                          group_by(SampleID, Warning, .drop = FALSE) %>% 
                          summarise(Warning_num=n())), by=c('SampleID')) %>% 
                          filter(Warning == "WARN") %>% 
                          mutate(Warning_percent = Warning_num/N) %>% 
                          select(SampleID, Warning_percent) %>% 
                          as.data.frame() 




BioMe_sample_LOD_data<-  BioMe_proteome_protname  %>% 
                          group_by(SampleID) %>% 
                          summarise(N = n()) %>% 
                          inner_join((BioMe_proteome_protname %>% 
                          group_by(SampleID, LOD_ind, .drop = FALSE) %>% 
                          summarise(LOD_num=n())), by=c('SampleID')) %>% 
                          filter(LOD_ind == 1) %>% 
                          mutate(Percentage_above_LOD = LOD_num/N) %>% 
                          select(SampleID, Percentage_above_LOD) %>% 
                          as.data.frame() 

BioMe_sample_check_data<- BioMe_sample_QC_data %>% 
                            inner_join(BioMe_sample_LOD_data, by=c('SampleID'))


BioMe_sample_check_table<- BioMe_sample_check_data %>% 
                             reactable(filterable = TRUE,  
                                    columns = list(Warning_percent = colDef(format = colFormat(percent = TRUE, digits = 1),
                                                                                 style = function(Warning_percent) {
                                                                                         if (Warning_percent > 0.25) {
                                                                                             color <- "#e00000"
                                                                                         } else if (Warning_percent <= 0.25) {
                                                                                             color <- "#008000"
                                                                                         } 
                                                                                           list(color = color, fontWeight = "bold")
                                                                                         }),
                                                   Percentage_above_LOD = colDef(format = colFormat(percent = TRUE, digits = 1),
                                                                                 style = function(Percentage_above_LOD) {
                                                                                         if (Percentage_above_LOD > 0.5) {
                                                                                             color <- "#008000"
                                                                                         } else if (Percentage_above_LOD <= 0.5) {
                                                                                             color <- "#e00000"
                                                                                         } 
                                                                                           list(color = color, fontWeight = "bold")
                                                                                         })))



BioMe_sample_check_table

BioMe_sample_warn_num<- BioMe_sample_check_data %>% 
                           filter(Warning_percent > 0.5) %>%
                           select('SampleID', 'Warning_percent') %>% 
                           reactable(columns = list(
                                       Warning_percent = colDef(format = colFormat(percent = TRUE, digits = 1)),
                                       SampleID = colDef(footer = function(values) sprintf("Total = %d", length(values)))))
                           
BioMe_sample_warn_num                          



BioMe_sample_LOD_num<- BioMe_sample_check_data %>% 
                           filter(Percentage_above_LOD < 0.5) %>%
                           select('SampleID', 'Percentage_above_LOD') %>% 
                           reactable(columns = list(
                                       Percentage_above_LOD = colDef(format = colFormat(percent = TRUE, digits = 1)),
                                       SampleID = colDef(footer = function(values) sprintf("Total = %d", length(values)))))
                           
BioMe_sample_LOD_num  


```

### Total proteins & samples
```{r, warning=FALSE,message=FALSE}
## pattern
BioMe_proteome_protname_QC<- BioMe_proteome_protname %>% 
                             mutate(value = if_else(Warning == "PASS", NPX, NA))

BioMe_proteome_protname_QC_wide<- BioMe_proteome_protname_QC %>% 
                      select(SampleID, OlinkID, value) %>% 
                      pivot_wider(
                        names_from = OlinkID,
                        values_from = value
                      )

vis_miss(BioMe_proteome_protname_QC_wide, warn_large_data = FALSE) +
  labs(title="QC_Warning")



BioMe_proteome_protname_LOD<- BioMe_proteome_protname %>% 
                             mutate(value = if_else(LOD_ind == 1, NPX, NA))

BioMe_proteome_protname_LOD_wide<- BioMe_proteome_protname_LOD %>% 
                      select(SampleID, OlinkID, value) %>% 
                      pivot_wider(
                        names_from = OlinkID,
                        values_from = value
                      )

vis_miss(BioMe_proteome_protname_LOD_wide, warn_large_data = FALSE) +
  labs(title="LOD_Warning")



BioMe_proteome_protname_NA<- BioMe_proteome_protname %>% 
                             mutate(value = if_else(Warning == "PASS" & LOD_ind == 1, NPX, NA))

BioMe_proteome_protname_NA_wide<- BioMe_proteome_protname_NA %>% 
                      select(SampleID, OlinkID, value) %>% 
                      pivot_wider(
                        names_from = OlinkID,
                        values_from = value
                      )

vis_miss(BioMe_proteome_protname_NA_wide, warn_large_data = FALSE) +
  labs(title="QC_Warning & LOD_Warning")




## number
BioMe_proteome_totalcheck_num<- BioMe_proteome_check_data %>% 
                           filter(Warning_percent > 0.5 | Percentage_above_LOD < 0.5) %>%
                           select('Panel', 'Protein_name', 'Gene_name', 'OlinkID', 'Warning_percent', 'Percentage_above_LOD') %>% 
                           reactable(groupBy = c("Panel"),
                                     columns = list(
                                       Warning_percent = colDef(format = colFormat(percent = TRUE, digits = 1)),
                                       Percentage_above_LOD = colDef(format = colFormat(percent = TRUE, digits = 1)),
                                       Protein_name = colDef(footer = function(values) sprintf("Total = %d", length(values)))))
                           
BioMe_proteome_totalcheck_num  



BioMe_sample_totalcheck_num<- BioMe_sample_check_data %>% 
                           filter(Warning_percent > 0.5 | Percentage_above_LOD < 0.5) %>%
                           select('SampleID', 'Warning_percent', 'Percentage_above_LOD') %>% 
                           reactable(columns = list(
                                       Warning_percent = colDef(format = colFormat(percent = TRUE, digits = 1)),
                                       Percentage_above_LOD = colDef(format = colFormat(percent = TRUE, digits = 1)),
                                       SampleID = colDef(footer = function(values) sprintf("Total = %d", length(values)))))
                           
BioMe_sample_totalcheck_num  

```







### Distribution
```{r, warning=FALSE,message=FALSE}
BioMe_proteome_protname_QC_high<- highlight_key((BioMe_proteome_protname_QC %>% filter(Panel=="Cardiometabolic")), ~OlinkID)
BioMe_proteome_protname_QC_plot<- ggplot(BioMe_proteome_protname_QC_high) +
                               geom_density(aes(x = value, fill=OlinkID), alpha=0.3)+
                               theme(legend.position = NULL)

p<- BioMe_proteome_protname_QC_plot %>%
    ggplotly(
    hoverinfo = "x")%>%
    highlight(on = "plotly_hover", off = "plotly_doubleclick")

htmlwidgets::saveWidget(p, "distribution_plot.html", selfcontained = F, libdir = "lib")
```








## Dealing duplicates
### Indoleamine 2,3-dioxygenase 1
Neurology_II & Oncology_II
```{r, warning=FALSE,message=FALSE}

cor(BioMe_proteome_wide$OID31050,  BioMe_proteome_wide$OID31474)

```

### Interleukin-6
Inflammation

### Interleukin-8
Inflammation & Oncology
```{r, warning=FALSE,message=FALSE}

cor(BioMe_proteome_wide$OID20631,  BioMe_proteome_wide$OID21430)

```

### Leiomodin-1
Neurology_II & Oncology_II
```{r, warning=FALSE,message=FALSE}

cor(BioMe_proteome_wide$OID31014,  BioMe_proteome_wide$OID31416)

```


### Protein scribble homolog
Neurology_II & Oncology_II
```{r, warning=FALSE,message=FALSE}

cor(BioMe_proteome_wide$OID31046,  BioMe_proteome_wide$OID31455)

```

### Tumor necrosis factor
Cardiometabolic & Neurology & Oncology
```{r, warning=FALSE,message=FALSE}

cor(BioMe_proteome_wide[, c('OID20074', 'OID20848', 'OID21237')])

```

<!-- # Heatmap -->
<!-- ```{r, warning=FALSE,message=FALSE} -->
<!-- first15samples <- BioMe_proteome_protname %>%  -->
<!--                   pull(SampleID) %>%  -->
<!--                   unique() %>%  -->
<!--                   head(15) -->
<!-- first10protein <- BioMe_proteome_protname %>% -->
<!--                   filter(Panel=="Cardiometabolic") %>%  -->
<!--                   pull(OlinkID) %>%  -->
<!--                   unique() %>%  -->
<!--                   head(10) -->


<!-- Heatmap<- BioMe_proteome_protname %>%  -->
<!--           filter(Panel=="Cardiometabolic" & SampleID %in% first15samples & OlinkID %in% first10protein) %>%  -->
<!--           olink_heatmap_plot() -->

<!-- Heatmap -->
<!-- ``` -->

# Protein used in analysis
```{r, warning=FALSE,message=FALSE}
protein_in_panel<- fread("~/Projects/BioMe/proteome/input/analysis_sample/protein_in_panel.txt")

table(protein_in_panel$Panel)


```
















